Monte Carlo Synthetic Acceleration Solutions for the Simplified $P_N$ Approximation

### 2014-01-29 doc/ttsp2013/ttsp_2013.tex
adding mc theory to the ttsp paper

\documentclass[letterpaper,12pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tmadd,tmath}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{tmath,tmadd}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[mathcal]{euscript} 
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
\\ \href{mailto:sslattery@wisc.edu}{\texttt{sslattery@wisc.edu}}
}

@@

In this paper, we present Monte Carlo Synthetic Acceleration (MCSA) as
a linear solver technique for the simplified $P_N$ ($SP_N$)
discretization of the Boltzmann neutron transport equation. Using the
fully-formed linear operator for the transport problem, we explore
solutions to the $SP_N$ equations with MCSA using a challenging light
water reactor fuel assembly criticality calculation as the driving
problem. Due to neutron scattering in the light water moderator,
several difficulties arise when applying MCSA to these problems that
were not observed in previous work where the method was applied to
more opaque thermal radiaton transport systems. In order to
effectively solve the $SP_N$ equations for the model problem, an
algebraic preconditioning strategy is developed for MCSA where we
utilize incomplete factorizations as preconditioners to achieve
convergence.  Using the preconditioners, MCSA performance is compared
with production impelementations of the preconditioned GMRES and
BiCGSTAB methods. Although the algebraic preconditioning strategy
presented in this work will not be practical from a high performance
implementation standpoint with results demonstrating poor time to
solution, it does indicate that the iterative performance of the
method is competitive with the preconditioned Krylov methods when
operator spectrum is properly conditioned.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization, and high
confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems are often variants on
the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
fission reactor neutronics simulations, the $S_N$ method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
an ad-hoc, multidimensional extension of the simple single dimension
planar $P_N$ equations that created a system of coupled,
diffusion-like equations known as the simplified $P_N$ ($SP_N$)
equations (cite gelbard). The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required.

As machines begin to operate at hundreds of petaflops peak performance
and beyond, trends toward reduced energy consumption will require
incredibly high levels of concurrency to achieve the desired
computation rates. The end result of these hardware changes is that
the larger number of low-powered processors will be prone to both soft
failures such as bit errors in floating point operations and hard
failures where the data owned by that processor cannot be
recovered. Because these failures are predicted to be common,
resilient solver technologies are required to overcome these events at
the application level. With solutions to deterministic transport
problems based on Monte Carlo techniques, such issues are potentially
alleviated by statistical arguments. In the case of soft failures,
isolated floating point errors in the Monte Carlo simulation are
absorbed within tally statistics while a hard failure is treated as a
high variance event where some portion of the Monte Carlo histories
lost.

Considering these physics and hardware issues, this goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. Using these numerical studies, we look
forward to parallelizing them with the goal that they may be leveraged
competitively on leadership-class computing platforms.

In previous work \cite{evans_monte_2012}, we applied the Monte Carlo
synthetic acceleration (MCSA) method the thermal radiation diffusion
equation. For several transient problems we observed that MCSA
performed competetively with preconditioned Krylov methods. In a
continuation of these studies, we have applied the method light water
reactor problems discretized by the $SP_N$ equations. This system has
no time dependence, a large scattering component, and assymetry in the
linear operator, creating a set of problems complementary to the
transient, opaque, and symmetric systems of our previous analyis. In
addition, solving the $SP_N$ equations in this way also breaks away
from the $S_N$ forms of parallelism where spatial parallelism is
achieved by an efficient parallel sweep, angular efficiency achieved
by pipe-lining, and energy parallelism achieved by decoupling the
groups. With the $SP_N$ equations as a full matrix system, we now can
parallelize the problem as prescribed by the Monte Carlo algorithm,
which may be significantly more scalable than current $S_N$ transport
practices.

@@

@@

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2012} for a more rigorous defintion of the method.

>>>>

\documentclass[letterpaper,11pt]{article}
\usepackage[top=1.0in,bottom=1.0in,left=1.0in,right=1.0in]{geometry}
\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{tmadd,tmath}
\usepackage{longtable}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{tmath,tmadd}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage[mathcal]{euscript} 
\usepackage[usenames]{color}
\usepackage[
naturalnames = true, 
colorlinks = true, 
linkcolor = black,
anchorcolor = black,
citecolor = black,
menucolor = black,
urlcolor = blue
]{hyperref}

%%---------------------------------------------------------------------------%%
\author{Stuart R. Slattery, Thomas M. Evans, Steven P. Hamilton
\\ \href{mailto:slatterysr@ornl.gov}{\texttt{slatterysr@ornl.gov}}
}

@@

We present Monte Carlo Synthetic Acceleration (MCSA) as a linear
solver technique for the simplified $P_N$ ($SP_N$) discretization of
the Boltzmann neutron transport equation. Using the fully-formed
linear operator for the transport problem, we explore solutions to the
$SP_N$ equations with MCSA using a challenging light water reactor
fuel assembly criticality calculation as the driving problem. Due to
neutron scattering in the light water moderator, several difficulties
arise when applying MCSA to these problems that were not observed in
previous work where the method was applied to more opaque thermal
radiaton transport systems. Of most significance, it was discovered
that simply reducing the spectral radius of the Monte Carlo problem
below unity is not enough to achieve convergence as outlined in
previous work. In order to effectively solve the $SP_N$ equations for
the model problem, an algebraic preconditioning strategy is developed
for MCSA where we utilize incomplete factorizations as preconditioners
to achieve convergence. Using the preconditioners, MCSA performance is
compared with production implementations of the preconditioned GMRES
and BiCGSTAB methods. Studies indicate that the iterative performance
of the method is competitive with the preconditioned Krylov methods
when operator spectrum is properly conditioned.

@@

Predictive modeling and simulation capability requires the combination
of high fidelity models, high performance computing hardware that can
handle the intense computational loads required by these models, and
modern algorithms for solving these problems that leverage this high
performance hardware. For nuclear reactor analysis, this predictive
capability can enable tighter design tolerances for improved thermal
performance and efficiency, higher fuel utilization, and high
confidence in accident scenario models. Modern high fidelity
deterministic methods for large scale problems are often variants on
the discrete ordinates ($S_N$) method \cite{evans_denovo:_2010}. For
fission reactor neutronics simulations, the $S_N$ method requires
potentially trillions of unknown angular flux moments to be computed
to achieve good accuracy for the responses of interest
\cite{slaybaugh_acceleration_2011}. In the 1960's, Gelbard developed
an ad-hoc, multidimensional extension of the simple single dimension
planar $P_N$ equations that created a system of coupled,
diffusion-like equations known as the simplified $P_N$ ($SP_N$)
equations (cite gelbard). The reduction in numerical complexity
compared to current deterministic solution methods using the $S_N$
approximation could mean significant savings in both compute time and
memory required and permit larger simulations to be performed on the
available computing platforms.

Looking towards future computing platforms, as machines begin to
operate at hundreds of petaflops peak performance and beyond, trends
toward reduced energy consumption will require incredibly high levels
of concurrency to achieve the desired computation rates. The end
result of these hardware changes is that the larger number of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events at the application level. With solutions to
deterministic transport problems based on Monte Carlo techniques, such
issues are potentially alleviated by statistical arguments. In the
case of soft failures, isolated floating point errors in the Monte
Carlo simulation are absorbed within tally statistics while a hard
failure is treated as a high variance event where some portion of the
Monte Carlo histories lost.

Considering these physics and hardware issues, the goal of this work
is to research and develop Monte Carlo Synthetic Acceleration methods
for neutron transport problems. These methods are based on using the
Monte Carlo method to directly accelerate the convergence of
traditional fixed-point iterative schemes for these problems. We aim
to address both their benefits and their shortcomings in the context
of the physics of interest and desire to identify areas in which
improvements can be made. In previous work \cite{evans_monte_2012}, we
applied the Monte Carlo synthetic acceleration (MCSA) method to the
thermal radiation diffusion equation. For several transient problems
we observed that MCSA performed competetively with preconditioned
Krylov methods. In a continuation of these studies, we apply the
method to light water reactor problems discretized by the $SP_N$
equations. This system has no time dependence, a large scattering
component, and assymetry in the linear operator, creating a set of
problems complementary to the transient, opaque, and symmetric systems
of our previous analyis. In addition, we apply a new fixed point
iteration in the MCSA sequence as well as a new estimator in order to
study their effects on convergence.

@@

\subsection{Monte Carlo Synthetic Acceleration}
\label{subsec:mcsa}
In this section we briefly review the Monte Carlo method for linear
systems and Monte Carlo Synthetic Acceleration. Given a linear
problem:
\begin{equation}
  \ve{A} \ve{x} = \ve{b}\:,
  \label{eq:linear_problem}
\end{equation}
where $\ve{A} \in \mathbb{R}^{N \times N}$ is a matrix operator such
that $\ve{A} : \mathbb{R}^{N} \rightarrow \mathbb{R}^{N}$, $\ve{x} \in
\mathbb{R}^N$ is the solution vector, and $\ve{b} \in
\mathbb{R}^N$. To form the Monte Carlo method we define the linear
system adjoint to Eq~(\ref{eq:linear_problem}):
\begin{equation}
  \ve{A}^T \ve{y} = \ve{d}\:,
  \label{eq:adjoint_linear_problem}
\end{equation}
where $\ve{y}$ and $\ve{d}$ are the adjoint solution and source
vectors respectively and $\ve{A}^T$ is the adjoint operator for
$\ve{A} \in \mathbb{R}^{N \times N}$ which we hten split to get:
\begin{equation}
  \ve{y} = \ve{H}^T \ve{y} + \ve{d}\:,
  \label{eq:adjoint_split_system}
\end{equation}
where $\ve{H}^T$ is the adjoint iteration matrix. The spectral radius
of $\ve{H}^T$ must remain less than 1 for convergence. By defining the
following adjoint inner product equivalence \cite{spanier_monte_1969}:
\begin{equation}
  \langle \ve{A}^T \ve{y}, \ve{x} \rangle = \langle \ve{y}, \ve{A}
  \ve{x} \rangle\:.
  \label{eq:adjoint_operator_product}
\end{equation}
it follows that:
\begin{equation}
  \langle \ve{x}, \ve{d} \rangle = \langle \ve{y}, \ve{b} \rangle\:.
  \label{eq:adjoint_vector_relation}
\end{equation}
We can acquire the adjoint solution by forming the Neumann series by
writing Eq~(\ref{eq:adjoint_split_system}) as:
\begin{equation}
  \ve{y} = (\ve{I} - \ve{H}^T)^{-1} \ve{d}\:,
  \label{eq:adjoint_split_system_2}
\end{equation}
which in turn yields the Neumann series using the adjoint operator:
\begin{equation}
  \ve{y} = \sum_{k=0}^{\infty} (\ve{H}^T)^k\ve{d}\:.
  \label{eq:adjoint_neumann_series}
\end{equation}
We expand this summation to again yield a series of transitions that
can be approximated by a Monte Carlo random walk sequence by forming
the Neumann series in reverse order:
\begin{equation}
  y_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i_k,i_{k-1}}\ldots h_{i_2,i_1} h_{i_1,i} d_{i_k}\:.
  \label{eq:adjoint_neumann_solution}
\end{equation}
We can readily build an estimator for the adjoint solution from this
series expansion, but we instead desire the solution to
Eq~(\ref{eq:linear_problem}). Here we have 2 unknowns, $\ve{y}$ and
$\ve{d}$, and therefore we require two constraints to close the
system. We use Eq~(\ref{eq:adjoint_vector_relation}) as the first
constraint and as a second constraint we select:
\begin{equation}
  \ve{d} = \boldsymbol{\delta}_j\:,
  \label{eq:adjoint_second_constraint}
\end{equation}
where $\boldsymbol{\delta}_j$ is one of a set of vectors in which the
$j^{th}$ component is the Kronecker delta function $\delta_{i,j}$. If
we apply Eq~(\ref{eq:adjoint_second_constraint}) to our first
constraint Eq~(\ref{eq:adjoint_vector_relation}), we get the following
convenient outcome:
\begin{equation}
  \langle \ve{y}, \ve{b} \rangle = \langle \ve{x},
  \boldsymbol{\delta}_j \rangle = x_j \:,
  \label{eq:inner_product_constraint}
\end{equation}
meaning that if we compute the inner product of the original source
and the adjoint solution using a delta function source, we recover one
component of the original solution.

In terms of radiation transport, this adjoint method is equivalent to
a traditional forward method where the initial state $i_0$ of the
random walk is determined by sampling the source vector $\ve{b}$ with
probabilities:
\begin{equation}
  P_{(i_0=i)}(\nu) = \frac{|b_i|}{||\ve{b}||_1}\:,
  \label{eq:adjoint_source_probability}
\end{equation}
with a random walk starting weight of:
\begin{equation}
  W_0 = ||\ve{b}||_1 \frac{b_i}{|b_i|}\:,
  \label{eq:adjoint_starting_weight}
\end{equation}
which gives the additional useful relation:
\begin{equation}
  b_{i_0} = W_0 P_{(i_0=i)}\:.
  \label{eq:adjoint_source_definition}
\end{equation}
We build weights and probabilities for the random walk using the
\textit{adjoint Neumann-Ulam decomposition} of $\ve{H}$:
\begin{equation}
  \ve{H}^{T} = \ve{P} \circ \ve{W}\:,
  \label{eq:adjoint_neumann_ulam}
\end{equation}
where we are forming the decomposition with respect to the transpose
of $\ve{H}$. $\ve{P}$ is defined element-wise as:
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_j |h_{ji}|}\:,
  \label{eq:adjoint_probability}
\end{equation}
such that we expect to select a new state, $j$, from the current state
in the random walk, $i$, by sampling column-wise (or row-wise if an
adjoint probability matrix is formed). Per
Eq~(\ref{eq:adjoint_neumann_ulam}), the transition weight is then
defined as:
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint_weight}
\end{equation}
Using the decomposition we can then define an expectation value for
the adjoint method. The contribution to the solution in state $j$ from
a particular random walk permutation of $k$ events is given by
\textit{collision estimator}:
\begin{equation}
  X_{j}(\nu) = \sum_{m=0}^k W_{m} \delta_{i_m,j}\:,
  \label{eq:adjoint_permutation_contribution}
\end{equation}
where the Kronecker delta indicates that the tally contributes only in
the current state, $i_m$, of the random walk. The expectation value
using all permutations is then:
\begin{equation}
  E\{X_j\} = \sum_{\nu} P_{\nu} X_{j}(\nu)\:,
  \label{eq:adjoint_expectation_value}
\end{equation}
which, if expanded utilizing Eq~(\ref{eq:adjoint_source_definition})
to insert the source term, directly recovers the exact solution:
\begin{equation}
  \begin{split}
    E\{X_j\} &=\sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N} b_{i_0} p_{i_0,i_1}p_{i_1,i_2}\ldots
    p_{i_{k-1},i_k} w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{k-1},i_k} \delta_{i_k,j} \\ &= x_{j}\:,
  \end{split}
  \label{eq:adjoint_expectation_expansion}
\end{equation}
therefore, also providing an unbiased Monte Carlo estimate of the
solution. It should be noted here that
Eq~(\ref{eq:adjoint_expectation_expansion}) only computes a single
component of our desired solution vector when really what we desire is
the entire solution vector. In an adjoint Monte Carlo simulation using
this estimator, the $w_{ij}$ elements that are added into the tally
for each state are only selected if/when the random walk currently
resides in that state. Much like a mesh tally in a particle transport
simulation, we have $N$ simultaneous tallies for $\ve{A} \in
\mathbb{R}^{N \times N}$ that will yield the entire solution vector.

In addition to the collision estimator, an additional estimator is
available due to the work of Okten \cite{okten_solving_2005} that uses
the method of expected values as a means to improve the Monte Carlo
estimate. As outlined by Spanier and Gelbard
\cite{spanier_monte_1969}, the method of expected values is a
deterministic averaging of events that may potentially occur in the
Monte Carlo random walk sequence. Okten applied this principle
directly to discrete Monte Carlo by forming the \textit{expected value
  estimator}\footnote{The variance for this estimator is discussed in
  Appendix~\ref{chap:estimator_variance}.} for a random walk of $k$
events:
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k W_m h_{j,i_m}\,
  \label{eq:expected_value_estimator}
\end{equation}
where now the contribution of the iteration matrix is
deterministically averaged at step $m$ over all potential states $j$
that may be reached from the current state $i_m$. Via Okten, the
estimator can be shown to be unbiased through a comparison to the
collision estimator. We can first rewrite the summation in
Eq~(\ref{eq:expected_value_estimator}):
\begin{equation}
  X_{j}(\nu) = b_j + \sum_{m=0}^k \sum_{i=1}^N W_m
  \delta_{i_m,i} h_{ji}\,
  \label{eq:unbiased_eval_1}
\end{equation}
where $N$ is the number of states in the system. Immediately, we see
the collision estimator as defined by
Eq~(\ref{eq:adjoint_permutation_contribution}) and can therefore write
the expectation value as:
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N E\{X_{i}\} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
which is equivalently is the $j^{th}$ component of
Eq~(\ref{eq:richardson_split}):
\begin{equation}
  E\{X_{j}\} = b_j + \sum_{i=1}^N x_{i} h_{ji}\,
  \label{eq:unbiased_eval_2}
\end{equation}
and is therefore an unbiased estimate. Compared to the collision
estimator, the expected value estimator provides additional
information at every step of the random walk, yielding potentially
better statistics with the same amount of transport
work. Conveniently, even if no Monte Carlo histories are computed, the
expected value estimator still deterministically computes the first
term of the Neumann Series, $\ve{H}^0\ve{b}$, whereas the collision
estimator will provide no information.

@@

At each eigenvalue iteration, we then solve the linear system with the
MCSA:
\begin{subequations}
  \begin{gather}
    \ve{r}_{l} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1/2} = \mathbf{\Phi}^{n+1}_l + \ve{r}_l\:,\\
    \ve{r}_{l+1/2} = \frac{1}{k^n}\ve{F}\mathbf{\Phi}^n -
    \ve{M}\mathbf{\Phi}^{n+1}_{l+1/2}\:,\\
    \ve{M}\delta\mathbf{\Phi}^{n+1}_{l+1/2} = \ve{r}_{l+1/2}\:,\\
    \mathbf{\Phi}^{n+1}_{l+1} = \mathbf{\Phi}^{n+1}_{l+1/2} + \delta
    \mathbf{\Phi}_{l+1/2}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where the $l$ subscript indicates the MCSA iteration number and $n$
again indicates the eigenvalue iteration. We refer the reader to
\cite{evans_monte_2012} for more a more complete definition of the
Monte Carlo method for linear systems and MCSA.
